Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2019-01-23 scagnostics.Rmd
Typo fixed.
Taking into account comments by German.

authors: 
- name: Ursula Laa
  thanks: The authors gratefully acknowledge the support of the Australian Research Council
  affiliation: Department of Physics, Monash University
  email: \email{ursula.laa@monash.edu}

@@

abstract: |
  Projection pursuit is used to find interesting low-dimensional projections of high-dimensional data by optimizing an index over all possible projections. Most indexes have been developed to detect departure from known distributions, such as normality, or to find separations between known groups. Here, we are interested in finding projections revealing potentially complex bivariate patterns, using new indexes constructed from scagnostics and a maximum information coefficient, with a purpose to detect unusual relationships between model parameters describing physics phenomena. The performance of these indexes is examined with respect to ideal behaviour, using simulated data, and then applied to several physics problems. The implementation builds upon the projection pursuit tools available in the R package, tourr, with indexes constructed from code in the R packages, scagnostics, minerva and mbgraphic.

@@

The motivation is based on physics applications, to aid the interpretation of model fits on experimental results. A physical model can be considered to be a set of $p$ free parameters, that cannot be measured directly and are determined by fitting a set of $q (p<q)$ experimental observations, for which predictions can be made once the $p$ parameters are estimated. (Note here, that while we often have analytic expressions for the predictions, this is not always the case and we may have to rely on numerical computation.) Different model fits compatible with the experimental results ($n$) yield the data to be examined using projection pursuit. A single prediction can be a complicated function of all of the free parameters, and typically $q \in [100,1000]$ and $p \sim 10$. Current practice is to examine pairs of parameters, or combinations produced by intuition or prior knowledge. This begs the question, whether important nonlinear associations are missed because tools to search are not available.

To find appropriate projections pursuit indexes, literature on variable selection was examined. With high-dimensional data, even plotting all pairs of variables can lead to many plots, which is what "scagnostics" (@scag, @WW08) were developed to address by providing metrics from which to select the most interesting variable pairs. There are eight scagnostics, of which three ("convex", "skinny" and "stringy") are used here. The question is whether these can be adapted into projection pursuit indexes, to search for unusual features in two-dimensional projections of high-dimensional data. Recent PhD research by @Grimm2016 explored the behaviour of scagnostics for selecting variables, and proposed two more that have nicer properties, based on smoothing splines and distance correlation. In addition, two more indexes for measuring dependence have been proposed in the machine learning literature, based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], with computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308]. These are related to original 1D projection pursuit indexes based on entropy (e.g. @huber85, @JS87). This provides seven current indexes for measuring dependence between two variables, and each is available in an R [@rref] package: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr], and provides optimisation routines, and visualization.

The paper investigates the behaviour of these indexes. Section \ref{sec:construct} discusses index construction, and how they fit into the guided tour. Section \ref{sec:investigate} investigates the behaviour of the indexes, explored primarily using tour methods (@As85, @BCAH05, @CBCH94). The new guided tour with these indexes are applied to two examples from physics models (Section \ref{sec:phys}).

@@

Making a plot always involves some choice of scaling. When a scatterplot is made, effectively, albeit under the hood, the data is scaled into a range of $[a, b]$ (often $a=0, b=1$) on both axes to print it on a page or display in a graphics device window. The range deliminates page space within which to draw. The upshot is that the original data scale is standardized to the range and aspect ratio on the display space. It maybe that the original range of one variable is $[1, 1000]$ and the other is $[1, 1.6]$ but the display linearly warps this to $[0, 1]$ and $[0, 1]$, say, giving both variables equal visual weight.

@@

- Standardizing each variable, to have mean 0 and variance 1, so that individual variable scales do not affect the result. Different variable scales are examinable without resorting to projection pursuit, so can be handled prior to searching through high dimensions.
- Sphering the high-dimensional data is often done to remove linear dependence. This is typically done using principal component analysis, and using the principal components as the variables passed to PP. If linear dependence is the only structure PP is not needed, and thus this is removed before PP so that the PPIs are not distracted by simple structure.
- Transform single variables to reduce skewness. It is marginal structure, visible in a single variable, which doesn't need a multivariate technique to reveal. Skewed distributions will inadvertently affect the PPIs, distracting the search for dependence. 
- Remove outliers, which may be an iterative process, to discover, identify and delete. Extreme values will likely affect PPI performance. Outliers can be examined on a case by case basis later.
- Possibly remove noise dimensions, which is also likely to be an iterative process. Directions where the distribution is purely noise make optimization of a PPI more difficult. If a variable is suspected to have little structure and relationship with other variables, conducting PP on the subset of variables without them may improve the efficiency of the search.
- Centering and scaling of the projected data, can be helpful visually. If the data has a small amount of non-normal distribution in some directions, the projected data can appear to wander around the plot window during a tour. It doesn't matter what the centre of the projected data is, so centering removes a wandering scatterplot. Less commonly, it may be useful to scale the projected data to standard values, which would be done to remove any linear dependence remaining in the data.

@@

Table \ref{notation} summarises the notation used for this section. This an overview of the functions that are converted into projection pursuit indexes. Full details of the functions can be found in the original sources.

@@

The structured projection is in variables 5 and 6 ($x_5, x_6$). Two samples sizes are used: $n=(100, 1000)$. All variables are standardized to have mean 0 and standard deviation 1. Figure \ref{fig:data} shows samples from each of the families, of the nuisance and structured pairs of variables. Table \ref{tab:indexTable} compares the PPIs for structured projections against those for nuisance variables, based on 100 simulated data sets of each type, using sample size 1000. The lower and upper show the 5th and 95th percentile of values. All indexes, except convex show distinctly higher values for the structured projections. The convex index shows the inverse scale to other indices, thus (1-convex) will be used in the assessment of performance of PPIs. The scale for the holes index in its original implementation is smaller than the others ranging from about 0.7 through 1, so it is re-scaled in the performance assessment so that all indices can be plotted on a common scale of 0-1 (details in Appendix). Similarly, the TIC index is re-scaled depending on sample size.

@@

```{r plotV1V2toV3V4, fig.width=6, fig.height=7, fig.cap="PPIs for projections along an interpolation between two nuisance projections. All projections would be nuisance so the PPI are ideally low and smooth, with little difference between sample sizes (solid lines: $n=1000$; dashed: $n=100$). The scagnostic PPIs are noisy. Some indexes have distinct differences in values between sample sizes."}
fullResMelt <- fullRes %>%
  mutate(convex = 1-convex) %>%
  rename("1-convex"=convex) %>%
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  gather(PPI, value, -t, -name, -size) %>%
  filter(PPI != "cmass") %>%
  mutate(PPI = fct_relevel(PPI, "holes", "cmass", "1-convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))

@@

```{r wIntermediate, fig.width=6, fig.height=7, fig.cap="PPIs for projections along an interpolation between nuisance and structured projections, following $x_1$-$x_2$ to $x_1$-$x_5$ to $x_5$-$x_6$ (solid lines: $n=1000$; dashed: $n=100$). Peaks at the end of the sequence indicate the index sees the structure. The scagnostics, MIC and TIC see all three structures, so are more flexible for general pattern detection. Holes only responds to the pipe, and is a multimodal function for this data with a local maximum at $x_1$-$x_5$. "}
intermediatePlane <- which(attributes(interpolate(t3))$new_basis)[2]
fullResMeltPlanned <- fullResPlanned %>%
  mutate(convex = 1-convex) %>%
  rename("1-convex"=convex) %>%
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  gather(PPI, value, -t, -name, -size) %>%
  filter(PPI != "cmass") %>%
  mutate(PPI = fct_relevel(PPI, "holes", "1-convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))
fullResMeltPlanned %>% 
  ggplot(aes(x=t, y=value)) + 
  geom_line(aes(group=size, linetype=factor(size, levels=c("1000", "100")))) + 
  ylim(c(0,1)) +
  geom_vline(xintercept=intermediatePlane, color="blue", alpha=0.3) +
  facet_grid(PPI~name) + 
  theme(legend.position="none") + xlab("Sequence of projections (t)") + ylab("PPI value")
```

@@

# Application to physics data
\label{sec:phys}

@@

With only five parameters, a reasonable start is to examine the 5D space using a grand tour. This quickly shows that the strong nonlinear relationships between the parameters. PP is then used to extract these relationships. The best index for this sort of problem is the splines2d, and it is fast to compute.

@@

This data contains posterior samples for simulation from a model describing a binary black hole (BBH) merger event. There are twelve model parameters. Flat priors are used for most model parameters.

@@

All seven PP indexes are applied to the data. Figure \ref{fig:bbhGuided} showing the projections that maximize three of the indexes. TIC and splines2d indexes identify very similar projections, that are based on the three parameters, dec, time and ra. This is to be expected based on the pairwise scatterplots. On the other hand, the 1-convex index finds a very different view, but actually the problem is that the optimization doesn't adequately reach a maximum for this index.

@@

# Extimating the squint angle of the spiral

>>>>

authors: 
- name: Ursula Laa
  thanks: The authors gratefully acknowledge the support of the Australian Research Council
  affiliation: Department of Physics, Monash University
  email: ursula.laa@monash.edu

@@

abstract: |
  Projection pursuit is used to find interesting low-dimensional projections of high-dimensional data by optimizing an index over all possible projections. Most indexes have been developed to detect departure from known distributions, such as normality, or to find separations between known groups. Here, we are interested in finding projections revealing potentially complex bivariate patterns, using new indexes constructed from scagnostics and a maximum information coefficient, with a purpose to detect unusual relationships between model parameters describing physics phenomena. The performance of these indexes is examined with respect to ideal behaviour, using simulated data, and then applied to problems from gravitational wave astronomy. The implementation builds upon the projection pursuit tools available in the R package, tourr, with indexes constructed from code in the R packages, scagnostics, minerva and mbgraphic.

@@

The motivation is based on physics applications, to aid the interpretation of model fits on experimental results. A physical model can be considered to be a set of $p$ free parameters, that cannot be measured directly and are determined by fitting a set of $q (p<q)$ experimental observations, for which predictions can be made once the $p$ parameters are estimated. (Note here, that while we may have analytic expressions for the predictions, this is not always the case and we often have to rely on numerical computation.) Different sets of model parameters ($n$) found to be  compatible with the experimental results within a selected level of confidence yield the data to be examined using projection pursuit. A single prediction can be a complicated function of all of the free parameters, and typically $q \in [100,1000]$ and $p \sim 10$. Current practice is to examine pairs of parameters, or combinations produced by intuition or prior knowledge. This begs the question, whether important nonlinear associations are missed because tools to search are not available.

To find appropriate projections pursuit indexes, literature on variable selection was examined. With high-dimensional data, even plotting all pairs of variables can lead to too many plots, which is what "scagnostics" (@scag, @WW08) were developed to address by providing metrics from which to select the most interesting variable pairs. There are eight scagnostics, of which three ("convex", "skinny" and "stringy") are used here. The question is whether these can be adapted into projection pursuit indexes, to search for unusual features in two-dimensional projections of high-dimensional data. Recent PhD research by @Grimm2016 explored the behaviour of scagnostics for selecting variables, and proposed two more that have nicer properties, based on smoothing splines and distance correlation. In addition, two more indexes for measuring dependence have been proposed in the machine learning literature, based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], with computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308]. These are related to original 1D projection pursuit indexes based on entropy (e.g. @huber85, @JS87). This provides seven current indexes for measuring dependence between two variables, and each is available in an R [@rref] package: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr], and provides optimisation routines, and visualization.

This paper investigates the behaviour of these indexes. Section \ref{sec:construct} discusses index construction, and how they fit into the guided tour. Section \ref{sec:investigate} investigates the behaviour of the indexes, explored primarily using tour methods (@As85, @BCAH05, @CBCH94). The new guided tour with these indexes is applied to two examples from gravitational wave astronomy (Section \ref{sec:phys}).

@@

Making a plot always involves some choice of scaling. When a scatterplot is made, effectively, albeit under the hood, the data is scaled into a range of $[a, b]$ (often $a=0, b=1$) on both axes to print it on a page or display in a graphics device window. The range deliminates page space within which to draw. The upshot is that the original data scale is standardized to the range and aspect ratio on the display space. It may be that the original range of one variable is $[1, 1000]$ and the other is $[1, 1.6]$ but the display linearly warps this to $[0, 1]$ and $[0, 1]$, say, giving both variables equal visual weight.

@@

- Standardizing each variable, to have mean 0 and variance 1, so that individual variable scales do not affect the result. Different variable scales are examinable without resorting to projection pursuit, so can be handled prior to searching through high dimensions.
- Sphering the high-dimensional data is often done to remove linear dependence. This is typically done using principal component analysis, and using the principal components as the variables passed to PP. If linear dependence is the only structure PP is not needed, and thus this is removed before PP so that the PPIs are not distracted by simple structure.
- Transform single variables to reduce skewness. It is marginal structure, visible in a single variable, which doesn't need a multivariate technique to reveal. Skewed distributions will inadvertently affect the PPIs, distracting the search for dependence. 
- Remove outliers, which may be an iterative process, to discover, identify and delete. Extreme values will likely affect PPI performance. Outliers can be examined on a case by case basis later.
- Possibly remove noise dimensions, which is also likely to be an iterative process. Directions where the distribution is purely noise make optimization of a PPI more difficult. If a variable is suspected to have little structure and relationship with other variables, conducting PP on the subset of variables without them may improve the efficiency of the search.
- Centring and scaling of the projected data, can be helpful visually. If the data has a small amount of non-normal distribution in some directions, the projected data can appear to wander around the plot window during a tour. It doesn't matter what the centre of the projected data is, so centring removes a wandering scatterplot. Less commonly, it may be useful to scale the projected data to standard values, which would be done to remove any linear dependence remaining in the data.

@@

Table \ref{notation} summarises the notation used for this section. Here we give an overview of the functions that are converted into projection pursuit indexes. Full details of the functions can be found in the original sources.

@@

The structured projection is in variables 5 and 6 ($x_5, x_6$). Two samples sizes are used: $n=(100, 1000)$. All variables are standardized to have mean 0 and standard deviation 1. Figure \ref{fig:data} shows samples from each of the families, of the nuisance and structured pairs of variables. Table \ref{tab:indexTable} compares the PPIs for structured projections against those for nuisance variables, based on 100 simulated data sets of each type, using sample size 1000. The lower and upper show the 5th and 95th percentile of values. The holes index is sensitive only to the pipe distribution. All other indexes, except convex show distinctly higher values for the structured projections. The convex index shows the inverse scale to other indices, thus (1-convex) will be used in the assessment of performance of PPIs. The scale for the holes index in its original implementation is smaller than the others ranging from about 0.7 through 1, so it is re-scaled in the performance assessment so that all indices can be plotted on a common scale of 0-1 (details in Appendix). Similarly, the TIC index is re-scaled depending on sample size.

@@

```{r plotV1V2toV3V4, fig.width=6, fig.height=7, fig.cap="PPIs for projections along an interpolation between two nuisance projections. All projections would be nuisance so the PPI are ideally low and smooth, with little difference between sample sizes (solid lines: $n=1000$; dashed: $n=100$). The scagnostic PPIs are noisy. Some indexes have distinct differences in values between sample sizes."}
namesFixed <- str_replace_all(fullRes$name, "sin", "sine")
fullResMelt <- fullRes %>%
  mutate(name = namesFixed) %>%
  mutate(convex = 1-convex) %>%
  rename("1-convex"=convex) %>%
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  gather(PPI, value, -t, -name, -size) %>%
  filter(PPI != "cmass") %>%
  mutate(PPI = fct_relevel(PPI, "holes", "cmass", "1-convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))

@@

```{r wIntermediate, fig.width=6, fig.height=7, fig.cap="PPIs for projections along an interpolation between nuisance and structured projections, following $x_1$-$x_2$ to $x_1$-$x_5$ to $x_5$-$x_6$ (solid lines: $n=1000$; dashed: $n=100$). Peaks at the end of the sequence indicate the index sees the structure. The scagnostics, MIC and TIC see all three structures, so are more flexible for general pattern detection. Holes only responds to the pipe, and is a multimodal function for this data with a local maximum at $x_1$-$x_5$. "}
intermediatePlane <- which(attributes(interpolate(t3))$new_basis)[2]
namesFixed <- str_replace_all(fullResPlanned$name, "sin", "sine")
fullResMeltPlanned <- fullResPlanned %>%
  mutate(name = namesFixed) %>%
  mutate(convex = 1-convex) %>%
  rename("1-convex"=convex) %>%
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  gather(PPI, value, -t, -name, -size) %>%
  filter(PPI != "cmass") %>%
  mutate(PPI = fct_relevel(PPI, "holes", "1-convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))
fullResMeltPlanned %>% 
  ggplot(aes(x=t, y=value)) + 
  geom_line(aes(group=size, linetype=factor(size, levels=c("1000", "100")))) + 
  ylim(c(0,1)) +
  geom_vline(xintercept=intermediatePlane, color="blue", alpha=0.3) +
  facet_grid(PPI~name) + 
  theme(legend.position="none") + xlab("Sequence of projections (t)") + ylab("PPI value")
```

@@

# Application to physics examples
\label{sec:phys}

@@

With only five parameters, a reasonable start is to examine the 5D space using a grand tour. This quickly shows  the strong nonlinear relationships between the parameters. PP is then used to extract these relationships. The best index for this sort of problem is the splines2d, and it is fast to compute.

@@

This data contains posterior samples from simulation from a model describing a binary black hole (BBH) merger event. There are twelve model parameters. Flat priors are used for most model parameters.

@@

All seven PP indexes are applied to the data. Figure \ref{fig:bbhGuided} showing the projections that maximize three of the indexes. TIC and splines2d indexes identify very similar projections, that are based on the three parameters, dec, time and ra. This is to be expected based on the pairwise scatterplots. On the other hand, the 1-convex index finds a very different view, but this is because the optimization doesn't adequately reach a maximum for this index.

@@

# Estimating the squint angle of the spiral
